\documentclass{tufte-book}

\hypersetup{colorlinks}% uncomment this line if you prefer colored hyperlinks (e.g., for onscreen viewing)

%%
% Book metadata
\title{Thesis \thanks{Thanks to Trevor Ba\vca, Josiah Wolf Oberholtzer, and Jeffrey Trevi\~no from whom I have learned so much.}}
\author[Gregory Rowland Evans]{Gregory Rowland Evans}


%%
% If they're installed, use Bergamo and Chantilly from www.fontsite.com.
% They're clones of Bembo and Gill Sans, respectively.
%\IfFileExists{bergamo.sty}{\usepackage[osf]{bergamo}}{}% Bembo
%\IfFileExists{chantill.sty}{\usepackage{chantill}}{}% Gill Sans

%\usepackage{microtype}

%%
% Just some sample text
\usepackage{lipsum}

%%
% For nicely typeset tabular material
\usepackage{booktabs}

%%
% For graphics / images
\usepackage{graphicx}
\setkeys{Gin}{width=\linewidth,totalheight=\textheight,keepaspectratio}
\graphicspath{{graphics/}}

% The fancyvrb package lets us customize the formatting of verbatim
% environments.  We use a slightly smaller font.
\usepackage{fancyvrb}
\fvset{fontsize=\normalsize}

%%
% Prints argument within hanging parentheses (i.e., parentheses that take
% up no horizontal space).  Useful in tabular environments.
\newcommand{\hangp}[1]{\makebox[0pt][r]{(}#1\makebox[0pt][l]{)}}

%%
% Prints an asterisk that takes up no horizontal space.
% Useful in tabular environments.
\newcommand{\hangstar}{\makebox[0pt][l]{*}}

%%
% Prints a trailing space in a smart way.
\usepackage{xspace}

%%
% Some shortcuts for Tufte's book titles.  The lowercase commands will
% produce the initials of the book title in italics.  The all-caps commands
% will print out the full title of the book in italics.
\newcommand{\vdqi}{\textit{VDQI}\xspace}
\newcommand{\ei}{\textit{EI}\xspace}
\newcommand{\ve}{\textit{VE}\xspace}
\newcommand{\be}{\textit{BE}\xspace}
\newcommand{\VDQI}{\textit{The Visual Display of Quantitative Information}\xspace}
\newcommand{\EI}{\textit{Envisioning Information}\xspace}
\newcommand{\VE}{\textit{Visual Explanations}\xspace}
\newcommand{\BE}{\textit{Beautiful Evidence}\xspace}

\newcommand{\TL}{Tufte-\LaTeX\xspace}

% Prints the month name (e.g., January) and the year (e.g., 2008)
\newcommand{\monthyear}{%
  \ifcase\month\or January\or February\or March\or April\or May\or June\or
  July\or August\or September\or October\or November\or
  December\fi\space\number\year
}


% Prints an epigraph and speaker in sans serif, all-caps type.
\newcommand{\openepigraph}[2]{%
  %\sffamily\fontsize{14}{16}\selectfont
  \begin{fullwidth}
  \sffamily\large
  \begin{doublespace}
  \noindent\allcaps{#1}\\% epigraph
  \noindent\allcaps{#2}% author
  \end{doublespace}
  \end{fullwidth}
}

% Inserts a blank page
\newcommand{\blankpage}{\newpage\hbox{}\thispagestyle{empty}\newpage}

\usepackage{units}

% Typesets the font size, leading, and measure in the form of 10/12x26 pc.
\newcommand{\measure}[3]{#1/#2$\times$\unit[#3]{pc}}

% Macros for typesetting the documentation
\newcommand{\hlred}[1]{\textcolor{Maroon}{#1}}% prints in red
\newcommand{\hangleft}[1]{\makebox[0pt][r]{#1}}
\newcommand{\hairsp}{\hspace{1pt}}% hair space
\newcommand{\hquad}{\hskip0.5em\relax}% half quad space
\newcommand{\TODO}{\textcolor{red}{\bf TODO!}\xspace}
\newcommand{\na}{\quad--}% used in tables for N/A cells
\providecommand{\XeLaTeX}{X\lower.5ex\hbox{\kern-0.15em\reflectbox{E}}\kern-0.1em\LaTeX}
\newcommand{\tXeLaTeX}{\XeLaTeX\index{XeLaTeX@\protect\XeLaTeX}}
% \index{\texttt{\textbackslash xyz}@\hangleft{\texttt{\textbackslash}}\texttt{xyz}}
\newcommand{\tuftebs}{\symbol{'134}}% a backslash in tt type in OT1/T1
\newcommand{\doccmdnoindex}[2][]{\texttt{\tuftebs#2}}% command name -- adds backslash automatically (and doesn't add cmd to the index)
\newcommand{\doccmddef}[2][]{%
  \hlred{\texttt{\tuftebs#2}}\label{cmd:#2}%
  \ifthenelse{\isempty{#1}}%
    {% add the command to the index
      \index{#2 command@\protect\hangleft{\texttt{\tuftebs}}\texttt{#2}}% command name
    }%
    {% add the command and package to the index
      \index{#2 command@\protect\hangleft{\texttt{\tuftebs}}\texttt{#2} (\texttt{#1} package)}% command name
      \index{#1 package@\texttt{#1} package}\index{packages!#1@\texttt{#1}}% package name
    }%
}% command name -- adds backslash automatically
\newcommand{\doccmd}[2][]{%
  \texttt{\tuftebs#2}%
  \ifthenelse{\isempty{#1}}%
    {% add the command to the index
      \index{#2 command@\protect\hangleft{\texttt{\tuftebs}}\texttt{#2}}% command name
    }%
    {% add the command and package to the index
      \index{#2 command@\protect\hangleft{\texttt{\tuftebs}}\texttt{#2} (\texttt{#1} package)}% command name
      \index{#1 package@\texttt{#1} package}\index{packages!#1@\texttt{#1}}% package name
    }%
}% command name -- adds backslash automatically
\newcommand{\docopt}[1]{\ensuremath{\langle}\textrm{\textit{#1}}\ensuremath{\rangle}}% optional command argument
\newcommand{\docarg}[1]{\textrm{\textit{#1}}}% (required) command argument
\newenvironment{docspec}{\begin{quotation}\ttfamily\parskip0pt\parindent0pt\ignorespaces}{\end{quotation}}% command specification environment
\newcommand{\docenv}[1]{\texttt{#1}\index{#1 environment@\texttt{#1} environment}\index{environments!#1@\texttt{#1}}}% environment name
\newcommand{\docenvdef}[1]{\hlred{\texttt{#1}}\label{env:#1}\index{#1 environment@\texttt{#1} environment}\index{environments!#1@\texttt{#1}}}% environment name
\newcommand{\docpkg}[1]{\texttt{#1}\index{#1 package@\texttt{#1} package}\index{packages!#1@\texttt{#1}}}% package name
\newcommand{\doccls}[1]{\texttt{#1}}% document class name
\newcommand{\docclsopt}[1]{\texttt{#1}\index{#1 class option@\texttt{#1} class option}\index{class options!#1@\texttt{#1}}}% document class option name
\newcommand{\docclsoptdef}[1]{\hlred{\texttt{#1}}\label{clsopt:#1}\index{#1 class option@\texttt{#1} class option}\index{class options!#1@\texttt{#1}}}% document class option name defined
\newcommand{\docmsg}[2]{\bigskip\begin{fullwidth}\noindent\ttfamily#1\end{fullwidth}\medskip\par\noindent#2}
\newcommand{\docfilehook}[2]{\texttt{#1}\index{file hooks!#2}\index{#1@\texttt{#1}}}
\newcommand{\doccounter}[1]{\texttt{#1}\index{#1 counter@\texttt{#1} counter}}

% Generates the index
\usepackage{makeidx}
\makeindex

\begin{document}

% Front matter
\frontmatter

% r.1 blank page
\blankpage

% v.2 epigraphs
\newpage\thispagestyle{empty}
\openepigraph{%
The public is more familiar with bad design than good design.
It is, in effect, conditioned to prefer bad design, 
because that is what it lives with. 
The new becomes threatening, the old reassuring.
}{Paul Rand%, {\itshape Design, Form, and Chaos}
}
\vfill
\openepigraph{%
A designer knows that he has achieved perfection 
not when there is nothing left to add, 
but when there is nothing left to take away.
}{Antoine de Saint-Exup\'{e}ry}
\vfill
\openepigraph{%
\ldots the designer of a new system must not only be the implementor and the first 
large-scale user; the designer should also write the first user manual\ldots 
If I had not participated fully in all these activities, 
literally hundreds of improvements would never have been made, 
because I would never have thought of them or perceived 
why they were important.
}{Donald E. Knuth}


% r.3 full title page
\maketitle


% v.4 copyright page
\newpage
\begin{fullwidth}
~\vfill
\thispagestyle{empty}
\setlength{\parindent}{0pt}
\setlength{\parskip}{\baselineskip}
Copyright \copyright\ \the\year\ \thanklessauthor

\par\smallcaps{Published by \thanklesspublisher}

\par\smallcaps{tufte-latex.github.io/tufte-latex/}

\par Licensed under the Apache License, Version 2.0 (the ``License''); you may not
use this file except in compliance with the License. You may obtain a copy
of the License at \url{http://www.apache.org/licenses/LICENSE-2.0}. Unless
required by applicable law or agreed to in writing, software distributed
under the License is distributed on an \smallcaps{``AS IS'' BASIS, WITHOUT
WARRANTIES OR CONDITIONS OF ANY KIND}, either express or implied. See the
License for the specific language governing permissions and limitations
under the License.\index{license}

\par\textit{First printing, \monthyear}
\end{fullwidth}

% r.5 contents
\tableofcontents

\listoffigures

\listoftables

% r.7 dedication
\cleardoublepage
~\vfill
\begin{doublespace}
\noindent\fontsize{18}{22}\selectfont\itshape
\nohyphenation
Dedicated to my teachers, who have given me everything,{} 
and the creators of \mbox{Abjad}.
\end{doublespace}
\vfill
\vfill


% r.9 introduction
\cleardoublepage
\chapter*{Epigraph}

This an introduction statement...


%%
% Start the main matter (normal chapters)
\mainmatter


\chapter{Introduction}
\label{ch:introduction}


\newthought{Emphasized} words can affect how they are read.

\newthought{Again} and again...

\vspace{-7\baselineskip}\footnotetext{THIS IS A FOOTNOTE.}
\vspace{7\baselineskip}

\bigskip
Paragraph

Another Paragraph

Another Paragraph

Another Paragraph

\newthought{New} thought
\bigskip

\newthought{New} thought



\section{Emph}\label{sec:label}\index{index}
\index{fonts|see{typefaces}}

Paragraph

Paragraph

Paragraph

\section{Headings}\label{sec:headings1}\index{headings}

Paragraphs can have sidenotes\sidenote{These Appear On The Side} and the paragraph continues on...

\paragraph{Paragraph} Paragraph.

\section{Sections can have titles}

Paragraph


\chapter[Chapter I: Some Prerequisite Knowledge]{Chapter I: Some Prerequisite Knowledge}
\label{ch:prerequisites}

Paragraph

\section{Section-Name}\label{sec:section-name}
\subsection{Subsection-Heading}\label{sec:heading}\index{heading}
words...

Paragraphs can have citations \cite{citations} which are good for academic papers.

Paragraph

% let's start a new thought -- a new section
\newthought{New Thought},\cite{citation} paragraph.

\section{New Section}

Paragraph

\chapter[Chapter II: My Compositional Practice with the Ecosystem of Python, Abjad, and Lilypond]{Chapter II: My Compositional Practice with the Ecosystem of Python, Abjad, and Lilypond}
\label{ch:practice}

\section{My Methodology}

        In the preceding chapter we have seen some of the strengths and potential weaknesses of abjad and lilypond when compared with similar software, as well as some potential logical pitfalls when working with these programs. In my recent compositional work, I have begun to amalgamate a workflow out of the ecosystem of python, abjad, and lilypond, by learning from and embracing the idiosyncrasies of the software.(will I actually address all that this implies? Maybe simplify?) Occaisonaly I have found the need to tweak abjad's source code in order for it to perform functions that I desire, but more often than this, the composer will find the need to build tools to simplify the process of engraving. The use of these tools in tandem is advantageous for my work due to the beautiful flexibility of lilypond's notational algorithm(what does this mean though? Spacing?font? Be clear?) and abjad's [(simplification of formal structures in music engraving) or (clear model of music notation)]. In the end, the greatest strength of this ecosystem is its flexibility. Not only are abjad and lilypond both immensely rich in their features, but, due to their open source nature, their source code is accessible to the user for further modification. In my work, I often desire a structural rigor, where rhythms, pitches, performance techniques, and instrumentation (among an infinite number of other parameters), are balanced together by a plan or logic that gives meaning (rightness or wrongness) to potential musical realities (for a given piece...I'm not setting rules for the universe, just my isolated universe where I am god). Very often, a rigorous structure falls apart when constructed by hand (humans are prone to err, while computers don't make mistakes unless you tell them to), thus working with the python programming language allows for a consistency in formal rigor that might be otherwise unattainable (it also allows for the potential modeling of complex systems or algorithmic music). (Since lilypond draws and has a looser{?} model of notation than other software, it allows the composer to have greater graphic freedom, but since it has no GUI it can spend more memory power when calculating collisions. Also since it allows you to include scheme functions, you are able to affect other features like proportional spacing across an entire score instead of manually dragging note heads like in finale or sibelius.) (Abjad seems tailor-made for polyphony (unlike OM or PWGL which seem to be hell-bent on single-voice-abstraction). It has no restrictions in this regard...and since you have direct access to the python code {unlike in OM or PWGL} you can VERY EASILY use THE SAME MATERIAL across multiple instruments at different points in time.) (Python's object-oriented potential models musical thought more clearly than LISP)
		
In this chapter we will take a look at the compositional advantages of working with these programs such as how to formalize(automate?) potentially tedious tasks, the benefits of an iterative compositional workflow, possibilities for composing with algorithms or models.

\section{Section}
\label{sec:label}

\chapter{Chapter III: Compositions}
\label{ch:compositions}

paragraph

\section{section}

paragraph

\chapter{Conclusion}
\label{ch:conclusion}
\section{Section}
Paragraph

\paragraph{Heading}
...paragraph...


%%
% The back matter contains appendices, bibliographies, indices, glossaries, etc.







\backmatter

\bibliography{sample-handout}
\bibliographystyle{plainnat}


\printindex

\end{document}

